A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-12 pkg/colorspace/paper/Toolbox/article.Rmd
first draft of summary/discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@485 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
approach, utiliting the perceptually-based HCL (hue--chroma--luminance) color space
[@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
Thus, based on this color model the \pkg{colorspace} package can derive general and adaptable
strategies for color palettes; manipulate individual colors and color palettes;
and assess and visualize the properties of color palettes (beyond simple color swatches).
Specifically, \pkg{colorspace} provides three types of palettes based on the HCL model:

@@

The \pkg{colorspace} package ships with a wide range of predefined color palettes,
specified through suitable trajectories in the HCL (hue-chroma-luminance) color space.
A quick overview can be gained easily with the `hcl_palettes()` function (see Figure\ \ref{fig:hcl-palettes}):

@@

In particular, the package provides various qualitative, sequential, and
diverging palettes derived by relatively simple trajectories in HCL
(Hue-Chroma-Luminance) space. In contrast to many other packages providing
modern balanced color palettes (such as \pkg{ColorBrewer.org}, \pkg{CARTO},
\pkg{viridis}, or \pkg{scico}) special emphasis is given to flexibility of
the palettes that can be adjusted to the particular needs of a given
data visulization. The manuscript already provided various tips and tricks
regarding factors that are relevant in choosing an effective palette
in a given situation. Further useful guidance is provided in many sources, including
@color:Ware:1988, @color:Okabe+Ito:2008, @color:Aigner:2010,
@color:Stauffer+Mayr+Dabernig:2015, @color:Zhang:2015, @color:Rost:2018,
@color:Wilke:2019, and @color:Ciechanowski:2019.

FIXME: Discuss relation ship to other packages:

* \pkg{Polychrome} [@color:Polychrome, @color:Polychrome2]: qualitative palettes with more "categories". This can generate 
  palettes with the (a large number of) 
  colours spaced far apart from each other in Luv space.
  It also has functions for visualising colour palettes in different ways 
  (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
* \pkg{roloc} [@color:roloc,@color:roloc2]: colour conversions, but not between
  numeric colour spaces;  instead, from numeric colour spaces to
  colour names.  
  (https://stattech.wordpress.fos.auckland.ac.nz/2018/01/25/2018-01-generating-colour-names-the-roloc-package-for-r/)
* \pkg{colorscience} [@color:colorscience]: This package is for Colour
Science.  It has data sets, does more complex conversions (including
different chromatic adaptation algorithms).  
So more sophisticated than \pkg{colorspace}
in terms of flexibility and low-level detail, BUT it does not do
palette-generation.
* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.

>>>>

The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
approach, utiliting the perceptually-based HCL (Hue--Chroma--Luminance) color space
[@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
Thus, based on this color model the \pkg{colorspace} package can derive general and adaptable
strategies for color palettes; manipulate individual colors and color palettes;
and assess and visualize the properties of color palettes (beyond simple color swatches).
Specifically, \pkg{colorspace} provides three types of palettes based on the HCL model:

@@

The \pkg{colorspace} package ships with a wide range of predefined color palettes,
specified through suitable trajectories in the HCL (Hue-Chroma-Luminance) color space.
A quick overview can be gained easily with the `hcl_palettes()` function (see Figure\ \ref{fig:hcl-palettes}):

@@

In particular, the package provides various qualitative, sequential, and
diverging palettes derived by relatively simple trajectories in HCL
(Hue--Chroma--Luminance) space. In contrast to many other packages providing
modern balanced color palettes (such as \pkg{ColorBrewer.org}, \pkg{CARTO},
\pkg{viridis}, or \pkg{scico}) special emphasis is given to flexibility of
the palettes that can be adjusted to the particular needs of a given
data visulization. The manuscript also provides various tips and tricks
for choosing an effective palette in a given situation. Further useful
guidance is provided in many sources, including:
@color:Ware:1988, @color:Okabe+Ito:2008, @color:Aigner:2010,
@color:Stauffer+Mayr+Dabernig:2015, @color:Zhang:2015, @color:Rost:2018,
@color:Wilke:2019, and @color:Ciechanowski:2019, among many others.

Further \proglang{R} packages that can complement the palettes provided
by \pkg{colorspace} include:
\pkg{Polychrome} [@color:Polychrome; @color:Polychrome2] implements strategies
for qualitative palettes with many "categories". While the qualitative palettes
in Section\ \ref{sec:hcl_palettes} yield only about 6--8 clearly distinguishable
colors due to the fixed chroma and luminance, \pkg{Polychrome} relaxes this
restriction and can thus find a larger number of colors in CIELUV space that
are spaced as far apart as possible. It also has several functions for visualizing
color palettes in different ways. 
The palette collection packages \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
also provide a wide range of prespecified palettes, including some qualitative
schemes with many categories. Note that the palettes are quite diverse, though,
and not all of them are equally suitable for coding qualitative information.
The visualization functions in \pkg{colorspace} from Section\ \ref{sec:palette_visualization}
may be helpful in assessing their properties.
\pkg{roloc} [@color:roloc; @color:roloc2] also provides color conversions, but
not between numeric colour spaces; instead, from numeric color spaces to
English color names.
